Nock for Everyday Coders

### 2025-08-28 mss.tex
Post
Post paginated version.
Add tombstone.
Remove hf
PosT

\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.10}
\usetikzlibrary{shapes.geometric,arrows,fit,matrix,positioning}
\tikzset
{
    treenode/.style = {circle, draw=black, align=center, minimum size=1cm},
    subtree/.style  = {circle, draw=darkgray, dashed, text=darkgray, align=center, minimum size=1cm}
}

@@

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} II:1 (2025):  1–46. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Nock for Everyday Coders}
\fancyfoot[LE,RO]{\thepage}

@@

\begin{lstlisting}[style=listingblock]
/[(a + a) b]        /[2 /[a b]]
\end{lstlisting}

The problem is, the programmer may already know that Nock code looks more like the below – just lists of numbers, with no symbols:

@@

Before getting started on Nock proper, we should understand how Nock and Hoon handle addresses in binary trees.  If you already understand why memory slot 5 of \lstinline[style=inlinecode]{[['apple' %pie] [0b1101 0xdad]]} is \lstinline[style=inlinecode]{%pie}, you are good to go and can skip ahead to the Section~\ref{sxn:nock0}.

@@

\begin{figure}[hbt!]
  \centering
  \caption{A binary tree with labeled node addresses to several layers. \label{fig:treefull}}
  \begin{tikzpicture}[
    ->, >=stealth,
    level 1/.style={sibling distance=8cm, level distance=1.5cm},
    level 2/.style={sibling distance=4cm, level distance=1.5cm},
    level 3/.style={sibling distance=2cm, level distance=1.5cm},
    level 4/.style={sibling distance=1cm, level distance=1.5cm},
    scale=0.6, transform shape,
    subtree/.style={circle, draw, fill=gray!20, minimum size=1cm},
    treenode/.style={circle, draw, fill=white, minimum size=1cm}
  ]
    \node [subtree] {\Large 1}
    child {
      node [subtree] {\Large 2}
      child {
        node [subtree] {\Large 4}
        child {
          node [subtree] {\Large 8}
          child { node [treenode] {\Large 16} }
          child { node [treenode] {\Large 17} }
        }
        child {
          node [subtree] {\Large 9}
          child { node [treenode] {\Large 18} }
          child { node [treenode] {\Large 19} }
        }
      }
      child {
        node [subtree] {\Large 5}
        child {
          node [subtree] {\Large 10}
          child { node [treenode] {\Large 20} }
          child { node [treenode] {\Large 21} }
        }
        child {
          node [subtree] {\Large 11}
          child { node [treenode] {\Large 22} }
          child { node [treenode] {\Large 23} }
        }
      }
    }
    child {
      node [subtree] {\Large 3}
      child {
        node [subtree] {\Large 6}
        child {
          node [subtree] {\Large 12}
          child { node [treenode] {\Large 24} }
          child { node [treenode] {\Large 25} }
        }
        child {
          node [subtree] {\Large 13}
          child { node [treenode] {\Large 26} }
          child { node [treenode] {\Large 27} } % Fixed duplicate 26 to 27
        }
      }
      child {
        node [subtree] {\Large 7}
        child {
          node [subtree] {\Large 14}
          child { node [treenode] {\Large 28} }
          child { node [treenode] {\Large 29} }
        }
        child {
          node [subtree] {\Large 15}
          child { node [treenode] {\Large 30} }
          child { node [treenode] {\Large 31} }
        }
      }
    };
  \end{tikzpicture}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \caption{A binary tree with some labeled node addresses. \label{fig:tree45}}
  {
    \begin{tikzpicture}
    [->,>=stealth',level/.style={sibling distance = 5cm/#1, level distance = 1.5cm},scale=0.6, transform shape]
    \node [subtree] {\Large 1}
    child {
      node [subtree] {\Large 2}
      child {
        node [treenode] {\Large 4}
      }
      child {
        node [treenode] {\Large 5}
      }
    }
    child {
      node [subtree] {\Large 3}
      child {
        node [treenode] {\Large 6}
      }
      child {
        node [subtree] {\Large 7}
        child {
          node [treenode] {\Large 14}
        }
        child {
          node [treenode] {\Large 15}
        }
      }
    }
    ;
    \end{tikzpicture}
  }
\end{figure}

@@

\begin{lstlisting}[style=listingcode]
:: get memory slot 1
> .*([50 51] [@0@ 1])
[50 51]
:: PSEUDOCODE -- replaces the Dojo's () with []
*[[50 51] [@0@ 1]]
:: *[a @0@ b] -> a = [50 51], b = 1
/[1 [50 51]]
[50 51]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*([50 51] [@0@ 2])
50
:: PSEUDOCODE -- replaces the Dojo's () with []
*[[50 51] [0 2]]
:: *[a @0@ b] -> a = [50 51], b = 2
/[2 [50 51]]
50
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*([20 30] [@1@ 67])
67
:: [@1@ 2 587] is same as [@1@ [2 [587]]]
> .*([20 30] [@1@ [2 [587])
[2 587]
\end{lstlisting}

@@

The answer is something that the Nock docs call the “distribution rule” or “implicit cons” (hello, fellow LISPers!), but that I find easiest to think of as the “Cell-Maker Rule”.

@@

Note that the pseudocode for \lstinline[style=inlinecode]{@2@} has nested `*`s.

@@

You now have seen all the fundamental functions/opcodes in Nock. In the next part, I'll introduce the remaining functions, which no longer need pseudocode: we have enough scaffolding now to build the rest of Nock from Nock itself. Instead, these new opcodes will just be shortcuts/macros/code expansions building on opcodes \lstinline[style=inlinecode]{@0@}-\lstinline[style=inlinecode]{@5@}.

@@

\noindent
\lstinline[style=inlinecode]{~/} ``fassig'' uses Nock \lstinline[style=inlinecode]{@11@} to pass the value \lstinline[style=inlinecode]{%my-function} to the interpreter, where it is associated with the C code that produces the product of \lstinline[style=inlinecode]{my-function}.

@@

In Hoon, you can acquire the Nock code for any function by using the \lstinline[style=inlinecode]{!=} ``zaptis'' rune.  Since this produces Nock code complete with gate calls, you may also find it helpful to cast a data type as a noun using \lstinline[style=inlinecode]{^-  *}, which will show you the raw Nock code for that value.

As you have seen, you can also evaluate raw Nock using the \lstinline[style=inlinecode]{.*} ``dottar'' rune, which evaluates to Nock opcode \lstinline[style=inlinecode]{@2@}.  Other Nock runes in Hoon include \lstinline[style=inlinecode]{.+}, which is opcode \lstinline[style=inlinecode]{@4@}; \lstinline[style=inlinecode]{.=}, which is opcode \lstinline[style=inlinecode]{@5@}; and \lstinline[style=inlinecode]{.?}, which is opcode \lstinline[style=inlinecode]{@3@}.  Fake opcode \lstinline[style=inlinecode]{@12@} is not actually Nock, but is used in a virtualization context to resolve values that may not be available in the current subject.

@@

::  now use that to set up the subject with a new tail (sample)
::  and call arm in memory slot 2 (our decrement formula)
> .*([dec-gate 36] [@9@ 2 @10@ [@3@ [@0@ 3]] @0@ 2])
35
> .*([dec-gate 562] [@9@ 2 @10@ [@3@ [@0@ 3]] @0@ 2])
561
\end{lstlisting}

@@

After this tutorial, you should be prepared to interpret Nock code when it is produced by Hoon.  Although there is only rarely any call to write Nock code directly, you should now have a good understanding of how to read and interpret it.  As you write your own interpreters in the future, you may even have call to hand-roll Nock expressions on occasion.

>>>>

\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.10}
\usetikzlibrary{shapes.geometric,arrows,fit,matrix,positioning}
\tikzset
{
    treenode/.style = {circle, draw=black, align=center, minimum size=1cm},
    subtree/.style  = {circle, draw=darkgray, dashed, text=darkgray, align=center, minimum size=1cm}
}
\renewcommand{\headrulewidth}{0pt}

@@

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} II:1 (2025):  1–45. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
% \fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
% \fancyhead[RO]{Nock for Everyday Coders}
% \fancyfoot[LE,RO]{\thepage}

@@

\begin{lstlisting}[style=listingcode]
/[(a + a) b]        /[2 /[a b]]
\end{lstlisting}

\noindent
The problem is, the programmer may already know that Nock code looks more like the below – just lists of numbers, with no symbols:

@@

Before getting started on Nock proper, we should understand how Nock and Hoon handle addresses in binary trees.  If you already understand why memory slot 5 of \lstinline[style=inlinecode]{[['apple' %pie] [0b1101 0xdad]]} is \lstinline[style=inlinecode]{%pie}, you are good to go and can skip ahead to Section~\ref{sxn:nock0}.

@@

\begin{figure}[hbt!]
  \centering
  \begin{tikzpicture}[
    ->, >=stealth,
    level 1/.style={sibling distance=8cm, level distance=1.5cm},
    level 2/.style={sibling distance=4cm, level distance=1.5cm},
    level 3/.style={sibling distance=2cm, level distance=1.5cm},
    level 4/.style={sibling distance=1cm, level distance=1.5cm},
    scale=0.6, transform shape,
    subtree/.style={circle, draw, fill=gray!20, minimum size=1cm},
    treenode/.style={circle, draw, fill=white, minimum size=1cm}
  ]
    \node [subtree] {\Large 1}
    child {
      node [subtree] {\Large 2}
      child {
        node [subtree] {\Large 4}
        child {
          node [subtree] {\Large 8}
          child { node [treenode] {\Large 16} }
          child { node [treenode] {\Large 17} }
        }
        child {
          node [subtree] {\Large 9}
          child { node [treenode] {\Large 18} }
          child { node [treenode] {\Large 19} }
        }
      }
      child {
        node [subtree] {\Large 5}
        child {
          node [subtree] {\Large 10}
          child { node [treenode] {\Large 20} }
          child { node [treenode] {\Large 21} }
        }
        child {
          node [subtree] {\Large 11}
          child { node [treenode] {\Large 22} }
          child { node [treenode] {\Large 23} }
        }
      }
    }
    child {
      node [subtree] {\Large 3}
      child {
        node [subtree] {\Large 6}
        child {
          node [subtree] {\Large 12}
          child { node [treenode] {\Large 24} }
          child { node [treenode] {\Large 25} }
        }
        child {
          node [subtree] {\Large 13}
          child { node [treenode] {\Large 26} }
          child { node [treenode] {\Large 27} } % Fixed duplicate 26 to 27
        }
      }
      child {
        node [subtree] {\Large 7}
        child {
          node [subtree] {\Large 14}
          child { node [treenode] {\Large 28} }
          child { node [treenode] {\Large 29} }
        }
        child {
          node [subtree] {\Large 15}
          child { node [treenode] {\Large 30} }
          child { node [treenode] {\Large 31} }
        }
      }
    };
  \end{tikzpicture}
  \caption{A binary tree with labeled node addresses to several layers. \label{fig:treefull}}
\end{figure}

@@

\begin{figure}[t]
  \centering
  {
    \begin{tikzpicture}
    [->,>=stealth',level/.style={sibling distance = 5cm/#1, level distance = 1.5cm},scale=0.6, transform shape]
    \node [subtree] {\Large 1}
    child {
      node [subtree] {\Large 2}
      child {
        node [treenode] {\Large 4}
      }
      child {
        node [treenode] {\Large 5}
      }
    }
    child {
      node [subtree] {\Large 3}
      child {
        node [treenode] {\Large 6}
      }
      child {
        node [subtree] {\Large 7}
        child {
          node [treenode] {\Large 14}
        }
        child {
          node [treenode] {\Large 15}
        }
      }
    }
    ;
    \end{tikzpicture}
  }
  \caption{A binary tree with some labeled node addresses. \label{fig:tree45}}
\end{figure}

@@

\begin{lstlisting}[style=listingcode]
:: get memory slot 1
> .*([50 51] [@0@ 1])
[50 51]
:: PSEUDOCODE -- replaces the Dojo's () with []
*[[50 51] [@0@ 1]]
:: *[a 0 b] -> a = [50 51], b = 1
/[1 [50 51]]
[50 51]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*([50 51] [@0@ 2])
50
:: PSEUDOCODE -- replaces the Dojo's () with []
*[[50 51] [0 2]]
:: *[a 0 b] -> a = [50 51], b = 2
/[2 [50 51]]
50
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*([20 30] [@1@ 67])
67
:: [1 2 587] is same as [1 [2 [587]]]
> .*([20 30] [@1@ [2 [587])
[2 587]
\end{lstlisting}

@@

The answer is something that the Nock docs call the “distribution rule” or “implicit cons” (hello, fellow \textsc{lisp}ers!), but that I find easiest to think of as the “Cell-Maker Rule”.

@@

Note that the pseudocode for \lstinline[style=inlinecode]{@2@} has nested \lstinline[style=inlinecode]{*}s.

@@

You now have seen all the fundamental functions/opcodes in Nock. In the next part, I'll introduce the remaining functions, which no longer need pseudocode: we have enough scaffolding now to build the rest of Nock from Nock itself. Instead, these new opcodes will just be shortcuts/macros/code expansions building on opcodes \lstinline[style=inlinecode]{@0@}--\lstinline[style=inlinecode]{@5@}.

@@

\noindent \sloppy
The Hoon rune \lstinline[style=inlinecode]{~/} fassig uses Nock \lstinline[style=inlinecode]{@11@} to pass the value \lstinline[style=inlinecode]{%my-function} to the interpreter, where it is associated with the C code that produces the product of \lstinline[style=inlinecode]{my-function}.

@@

In Hoon, you can acquire the Nock code for any function by using the \lstinline[style=inlinecode]{!=} zaptis rune.  Since this produces Nock code complete with gate calls, you may also find it helpful to cast a data type as a noun using \lstinline[style=inlinecode]{^-  *}, which will show you the raw Nock code for that value.

As you have seen, you can also evaluate raw Nock using the \lstinline[style=inlinecode]{.*} dottar rune, which evaluates to Nock opcode \lstinline[style=inlinecode]{@2@}.  Other Nock runes in Hoon include \lstinline[style=inlinecode]{.+} dotlus, which is opcode \lstinline[style=inlinecode]{@4@}; \lstinline[style=inlinecode]{.=} dottis, which is opcode \lstinline[style=inlinecode]{@5@}; and \lstinline[style=inlinecode]{.?} dotwut, which is opcode \lstinline[style=inlinecode]{@3@}.  Fake opcode \lstinline[style=inlinecode]{@12@} is not actually Nock, but is used in a virtualization context to resolve values that may not be available in the current subject.

@@

::  now use that to set up the subject with a new
::  tail (sample) and call arm in memory slot 2
::  (our decrement formula)
> .*([dec-gate 36] [@9@ 2 @10@ [@3@ [@0@ 3]] @0@ 2])
35
> .*([dec-gate 562] [@9@ 2 @10@ [@3@ [@0@ 3]] @0@ 2])
561
\end{lstlisting}

@@

After this tutorial, you should be prepared to interpret Nock code when it is produced by Hoon.  Although there is only rarely any call to write Nock code directly, you should now have a good understanding of how to read and interpret it.  As you write your own interpreters in the future, you may even have call to hand-roll Nock expressions on occasion. \tombstone
